Distributional Trees and Forests

### 2017-10-08 pkg/disttree/inst/draft/draft.Rnw
added plots to the rain section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1634 255f4249-2263-4417-b1ea-8498779e27a7

%% packages
\usepackage{amstext}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{thumbpdf}
\usepackage{rotating}
\usepackage{caption}
\captionsetup{format=hang}
\usepackage{subcaption}
\usepackage{Sweave}
\usepackage{tikz}
\usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}
%\usepackage[pagebackref=true]{hyperref}
%\usepackage{doi}

@@

\section{Introduction}

@@

All these parametric models work very well as long as only additive effects appear in the considered data. A different type of model that can capture non-additive effects as well are regression trees. Here the basic idea is to split the data set into more homogeneous subgroups based on information provided by the explanatory variables and to fit a model/response to each terminal node. One problem that might appear when applying tree algorithms is that they impose steps and abrupt changes on the model, even in situations where effects are rather smooth, as it can be seen in the top left plot of Figure \ref{Devel_treeforest}. One way how to tackle this problem is to not just consider one tree but an ensemble of trees. This is the idea of random forests. A set of trees is built, each on a slightly different data set which can either be a bootstrap or a subsample of the original data set. Then the results of all trees are combined to get a final model which unlike a tree model c!
 an also deal with smooth effects in a reasonable way.\\ 
In distributional trees and forests we now want to combine all these ideas from different fields of statistical modeling in order to benefit from all their advantages.\\
\\
In the past the combination of parametric data models and algorithmic tree models has gained more and more interest since it allows for a more flexible and clear structure. \\
Algorithms that only fit constants to each final node and therefore only model the mean of the response variable (such as CART, Breiman 1984) might lead to very large trees which complicates analyzing and interpreting. Algorithms such as GUIDE (Loh 2002), CRUISE (Kim and Loh 2001) and LOTUS (Chan and Loh 2004) were among the first to provide trees with parametric models in each terminal node. The MOB algorithm (Zeileis, Hothorn and Hornik 2008), an algorithm for model-based recursive partitioning, extended this idea by fitting parametric models to each node and using the gained information of the fitted model to create further splits in inner nodes. Using this algorithm as a framework to build trees LMs and GLMs can be fitted in the nodes of the trees. However, up to now trees which only model the mean of the response variable in its nodes are still more common and widely used.\\
Random forests (\cite{breiman2001random}) applying the CART algorithm to build the trees and cforest applying the ctree algorithm (\cite{hothorn2015partykit}, \cite{hothorn2006unbiased}) have used this idea of combining an ensemble of trees to get smoother effects of the covariates after combining or averaging over the parameters of the trees.\\
\\
But as mentioned before, until now the above listed development in the field of probabilistic modeling has not yet been fully integrated in the idea of algorithmic tree and forest models. Probabilistic models are rarely used in combination with tree algorithms and more flexible models such as GAMLSSs have not yet been applied as models in the nodes of a tree.\\
\\
Therefore, our main goal is to embed the progress made in the field of probabilistic modeling in the idea of regression trees and forests. In particular, more flexible models such as GAMLSSs should be fitted in the nodes of a tree in order to capture location, scale and shape as well as censoring, tail behavior etc. while non-additive effects or interactions of the explanatory variables can be detected automatically by the splitting algorithm used to build the tree. In that way, no variable selection or predefining of variable interactions is necessary and a whole distribution is specified in each node which leads to a wide range of inference methods that can be applied.\\
These (future) steps are illustrated in Figure \ref{Devel_treeforest}, embedded in the development made in the field of tree and forest algorithms.\\
\\
\\
Now we take a look at a classical text book example, the mcycle data set from the package \pkg{MASS}. It consists of a series of measurements of head acceleration in a simulated motorcycle accident. The response variable \code{accel} represents the recorded head acceleration meassured in g and the explanatory variable \code{times} is meassured in milliseconds after the impact.\\
We compare the performance of the following four functions on this data set: \code{disttree}, \code{distforest}, \code{gamlss} and \code{randomForest}\\
This situation is particularly well suited for a GAMLSS because there is only one regressor and physically one would expect a smooth function. However, this example shows, that the distributional forest can also capture the changes quite well albeit not as smoothly as the GAMLSS. Moreover, it illustrates that the distributional tree detects the changes as well but models them with rough steps.\\
\\
To make it possible to compare the models also on the estimated scale parameter, here the standard deviation of the random forest is estimated in each terminal node of each tree and then averaged, in the same way as the mean parameter is estimated. This makes it possible to specify a distribution, however, the splits of a random forest are only made in the location parameter while a distributional tree/forest considers splits in all distribution parameters.\\
\\
In Figure \ref{plot_mcycle} the results of a distributional forest are plotted on the top left and then compared to the results of a distributional tree, a GAMLSS and a random forest, regarding the location parameter on the top right and the scale parameter below.\\
While the smooth function of \code{gamlss} seems to model especially the middle part (for \code{times} between 17 and 40) better than the wiggly functions resulting from the trees it can also be observed that at \code{times} $=14$ there is a rapid drop which is captuerd more precisely by the tree-based models while \code{gamlss} starts descending earlier. Also for the highest values of \code{times} the smooth function increases again while the forests stay almost at the same level and the tree stays at a fixed value which seems more appropriate in this situation.\\
Comparing the two forest models, the distributional forest seems to be a bit more stable while the random forest includes more rapid changes of a wider range.\\
Looking at the plot of the scale parameter, in general the distributional forest estimates higher parameters than the random forest.\\

<<echo=FALSE, results=hide>>=

## FIX ME: distforest uses subsets, randomForest uses bootstraps

## randomForest: https://stackoverflow.com/questions/28417826/nodesize-parameter-ignored-in-randomforest-package
## nodesize = the minimum number of observations that must exist in a node in order for a split to be attempted.
## -> nodesize = minsplit

if(file.exists("mcycle_plotdata.rda")){
  load("mcycle_plotdata.rda")
} else {
  set.seed(7)
  dt <- disttree(accel ~ times, data = mcycle, family = NO(), type.tree = "mob", control = mob_control(minsize = 7L))
  df <- distforest(accel ~ times, data = mcycle, family = NO(), ntree = 300L, 
                   control = ctree_control(teststat = "quad", testtype = "Univ", mincriterion = 0, minbucket = 5L, minsplit = 10L, intersplit = TRUE))
  rf <- randomForest(accel ~ times, data = mcycle, ntree = 300L, nodesize = 10L, 
                     keep.inbag = TRUE, replace = FALSE)
  #ga <- gam(list(accel ~ s(times), ~ s(times)), data = mcycle, family=gaulss())
  ga <- gam(list(accel ~ s(times,k=20,bs="ad"), ~ s(times)), data = mcycle, family=gaulss())
  #b <- bamlss(list(accel ~ s(times), accel ~ s(times)), data = mcycle)
  #save(df, file = "mcycle_plotdata.rda")

  ## get 'fitted.sigma.rf'
  rf_sd <- rf_getsd(rf, newdata = mcycle, rfdata = mcycle)

  plotdata <- cbind(mcycle,
                    df$fitted.par,
                    dt$fitted.par,
                    #g$mu.fv, g$sigma.fv,
                    ga$fitted.values[,1], 1/ga$fitted.values[,2],
                    rf$predicted, rf_sd)

  colnames(plotdata) <- c("times", "accel",
                          "fitted.mu.df","fitted.sigma.df",
                          "fitted.mu.dt","fitted.sigma.dt",
                          "fitted.mu.g","fitted.sigma.g",
                          "fitted.mu.rf","fitted.sigma.rf")

  sp <- plotdata[order(plotdata["times"]),]

  save(sp, file = "mcycle_plotdata.rda")
}
@

\begin{figure}[t!]
\begin{subfigure}{0.65\textwidth}
<<plot_df, fig=TRUE, echo=FALSE>>=
# distforest
plot(mcycle)
lines(x = sp$times, y = sp$fitted.mu.df, type = "l", col = pal["forest"], lwd = 2)
polygon(c(sp$times, rev(sp$times)), c(sp$fitted.mu.df + sp$fitted.sigma.df, rev(sp$fitted.mu.df - sp$fitted.sigma.df)),
        col = pallight["forest"], border = "transparent")
@
%\caption{\label{plot_mcycle_df}Distributional Forest, \code{distforest}}
\end{subfigure}
\hspace{-2cm}
\begin{subfigure}{0.65\textwidth}
<<plot_mu_vs, fig=TRUE, echo=FALSE>>=
# disttree vs. distforest vs. gamlss vs. randomForest
plot(x = sp$times, y = sp$fitted.mu.dt, type = "l", col = pal["tree"], lwd = 1.5,
     ylim = c(-134,75), ylab = "accel", xlab = "times")
lines(x = sp$times, y = sp$fitted.mu.df, type = "l", col = pal["forest"], lwd = 1.5)
lines(x = sp$times, y = sp$fitted.mu.g, type = "l", col = pal["gamlss"], lwd = 1.5)
lines(x = sp$times, y = sp$fitted.mu.rf, type = "l", col = pal["randomForest"], lwd = 1.5)
legend('topleft', c("disttree", "distforest", "gamlss", "randomForest"), 
       col = c(pal["tree"], pal["forest"], pal["gamlss"], pal["randomForest"]), lty = 1, cex = 1)
@
%\caption{\label{plot_mcycle_mu_vs} Location parameter: \code{disttree} vs. \code{distforest} vs. \code{gamlss} vs. \code{randomForest}}
\end{subfigure}
\vspace{-1cm}

\hspace{4cm}
\begin{subfigure}{0.65\textwidth}
<<plot_sigma_vs_lines, fig=TRUE, echo=FALSE>>=
# variance: disttree vs. distforest vs. gamlss (lines)
plot(x = sp$times, y = sp$fitted.sigma.dt, type = "l", col = pal["tree"], lwd = 2,
     ylim = c(0,50), ylab = "variance of accel", xlab = "times")
#lines(x = sp$times, y = -sp$fitted.sigma.dt, type = "l", col = pal["tree"], lwd = 2)
lines(x = sp$times, y = sp$fitted.sigma.df, type = "l", col = pal["forest"], lwd = 2)
#lines(x = sp$times, y = -sp$fitted.sigma.df, type = "l", col = pal["forest"], lwd = 2)
lines(x = sp$times, y = sp$fitted.sigma.g, type = "l", col = pal["gamlss"], lwd = 2)
#lines(x = sp$times, y = -sp$fitted.sigma.g, type = "l", col = pal["gamlss"], lwd = 2)
lines(x = sp$times, y = sp$fitted.sigma.rf, type = "l", col = pal["randomForest"], lwd = 2)
#lines(x = sp$times, y = -sp$fitted.sigma.rf, type = "l", col = pal["randomForest"], lwd = 2)
legend('topleft', c("disttree", "distforest", "gamlss", "randomForest"), 
       col = c(pal["tree"], pal["forest"], pal["gamlss"], pal["randomForest"]), lty = 1, cex = 1)
@
\end{subfigure}
\caption{\label{plot_mcycle}top left: location and scale parameter of \code{distforest}\\
top right: location parameter of all four models\\
bottom: scale parameter of all four models}
\end{figure}

@@

\newpage
\section{Application}
\subsection{Continuation of the mcycle example: cross validation}
In the introduction the four functions \code{disttree}, \code{distforest}, \code{gamlss} and \code{randomForest} have been applied on the mcycle data set and their results haven been visualized and discussed. To get a more profound impression of their performances we now want to compare them in a cross validation framework on this text book example data set. The 133 observations are randomly split into 10 subsets roughly of the same size. After learning the model on 9 of these subgroups the RMSE of the estimated location parameter and the log-likelihood function are evaluated on the 10th part. Each subset is chosen once as a test data set and the mean of the resulting values is stored. This procedure is repeated 9 times. This comparison is illustrated in Figure \ref{plot_cross_mean}, based on the RMSE on the left and based on the log-likelihood on the right. (Extrapolation of the GAMLSS would lead to very small values of the log-likelihood. To avoid this, the first and the l!
 ast observation are always included in the learn data set.) \\
As we have already seen in the introduction the distributional tree obviously struggles with the smooth effects while the two forest models can capture the changes quite well. As far as the RMSE of the estimated location parameter is concerned the GAMLSS plays to its strength as a smooth parametric model and leads to the best results followed by the distributional forest and the random forest which might both be a bit to wiggly.\\
However, regarding the log-likelihood the forests perform better than the GAMLSS. This is mainly due to the fact that the GAMLSS models the scale parameter close to zero for observations with \code{times} $<10$. This leads to very low values of the log-likelihood function if one or more of these observations are included in the test data set. As the estimated scale parameter of the tree based models are higher for these specific observations this problem does not occure here.\\
Overall it can be seen that the distributional forest which fits the whole distribution and considers splits in both parameters performs better than the random forest which only models the mean and therefore also only inlcudes splits in the location parameter.

<<echo=FALSE, results=hide>>=
#### cross validation
if((file.exists("mcycle_cross_mean.rda")) & (file.exists("mcycle_cross_all.rda"))){
  load("mcycle_cross_mean.rda")
  load("mcycle_cross_all.rda")
  nrep_cross <- NROW(cross_all)
} else {

  set.seed(704)
  #set.seed(7054)
  nrep_cross <- 10

  rmse <- matrix(nrow = nrep_cross * 10, ncol = 4)
  ll <- matrix(nrow = nrep_cross * 10, ncol = 4)
  mean_rmse <- matrix(nrow = nrep_cross, ncol = 4)
  mean_ll <- matrix(nrow = nrep_cross, ncol = 4)
  errors <- numeric(length = nrep_cross * 10)
  colnames(rmse) <- colnames(mean_rmse) <- c("rmse_disttree", "rmse_distforest", "rmse_gamlss", "rmse_randomForest")
  colnames(ll) <- colnames(mean_ll) <- c("ll_disttree", "ll_distforest", "ll_gamlss", "ll_randomForest")

  tdata <- list()
  ldata <- list()

  for(k in 1:nrep_cross) {

    # separate data set (without the first and the last observation) into 10 parts (9 of size 13 and 1 of size 14):
    # (first and last observation are not in testid -> always stay in learndata)
    id <- c(2:(NROW(mcycle)-1))
    testid <- list()
    for(i in 1:9){
      testid[[i]] <- sample(id, 13, replace = FALSE)
      id <- id[!(id %in% testid[[i]])]
    }
    testid[[10]] <- id

    rmse_dt_mu <- rmse_df_mu <- rmse_g_mu <- rmse_rf_mu <- 
      ll_dt <- ll_df <- ll_g <- ll_rf <- numeric(length = 10)

    for(i in 1:10){
      tdata[[(k-1)*10+i]] <- testdata <- mcycle[testid[[i]], ]
      ldata[[(k-1)*10+i]] <- learndata <- mcycle[-testid[[i]], ]

      #testdata <- mcycle[testid[[i]], ]
      #learndata <- mcycle[-testid[[i]], ]

      dt <- disttree(accel ~ times, data = learndata, family = NO(), type.tree = "mob", control = mob_control(minsize = 7L))
      df <- distforest(accel ~ times, data = learndata, family = NO(), ntree = 300L, 
                       control = ctree_control(teststat = "quad", testtype = "Univ", mincriterion = 0, minbucket = 5L, minsplit = 10L, intersplit = TRUE))
      rf <- randomForest(accel ~ times, data = learndata, ntree = 300L, nodesize = 10L, keep.inbag = TRUE, replace = FALSE)
      ga <- gam(list(accel ~ s(times,k=20,bs="ad"), ~ s(times)), data = learndata, family=gaulss())
      #ga <- gam(list(accel ~ s(times), ~ s(times)), data = learndata, family=gaulss())
      errors[(k-1)*10+i] <- (ga$outer.info$conv != "full convergence")

      # get predicted values
      dt.pred.par <- predict(dt, newdata = testdata, type = "parameter")
      df.pred.par <- predict(df, newdata = testdata, type = "parameter")
      rf.pred.resp <- predict(rf, newdata = testdata)

      rf.sd <- rf_getsd(rf, newdata = testdata, rfdata = learndata)

      g.pred.par <- predict(ga, newdata = testdata, type = "response")
      g.pred.mu <- g.pred.par[,1]
      g.pred.sigma <- 1/g.pred.par[,2]

      # RMSE location parameter
      rmse_dt_mu[i] <- sqrt(mean((dt.pred.par$mu - testdata$accel)^2))
      rmse_df_mu[i] <- sqrt(mean((df.pred.par$mu - testdata$accel)^2))
      rmse_g_mu[i] <- sqrt(mean((g.pred.mu - testdata$accel)^2))
      rmse_rf_mu[i] <- sqrt(mean((rf.pred.resp - testdata$accel)^2))

      # loglikelihood
      dtll <- dfll <- gll <- rfll <- 0
      for(j in 1:(nrow(testdata))){

        dtll_j <- dnorm(testdata[j,"accel"], mean = dt.pred.par$mu[j], sd = dt.pred.par$sigma[j], log=TRUE)
        dfll_j <- dnorm(testdata[j,"accel"], mean = df.pred.par$mu[j], sd = df.pred.par$sigma[j], log=TRUE)
        gll_j <- dnorm(testdata[j,"accel"], mean = g.pred.mu[j], sd = g.pred.sigma[j], log=TRUE)
        rfll_j <- dnorm(testdata[j,"accel"], mean = rf.pred.resp[j], sd = rf.sd[j], log=TRUE)

        dtll <- if(is.na(dtll_j)) {
          print(dt.pred.par[j,], testdata[j,"accel"]) 
          dtll + (-5)
        } else {dtll + dtll_j}
        dfll <- if(is.na(dfll_j)) {
          print(df.pred.par[j,], testdata[j,"accel"]) 
          dfll + (-5)
        } else {dfll + dfll_j}
        gll <- if(is.na(gll_j)) {
          print(g.pred.mu[j], g.pred.sigma[j], testdata[j,"accel"]) 
          gll + (-5)
        } else {gll + gll_j}
        rfll <- if(is.na(rfll_j)) {
          print(c(rf.pred.resp[j], rf.sd[j]), testdata[j,"accel"]) 
          rfll + (-5)
        } else {rfll + rfll_j}
      }

      ll_dt[i] <- dtll
      ll_df[i] <- dfll
      ll_g[i] <- gll
      ll_rf[i] <- rfll

      rmse[i+(k-1)*10,] <- c(rmse_dt_mu[i], rmse_df_mu[i], rmse_g_mu[i], rmse_rf_mu[i])
      ll[i+(k-1)*10,] <- c(ll_dt[i], ll_df[i], ll_g[i], ll_rf[i])
    }

    # store data
    mean_rmse[k,] <- c(mean(rmse_dt_mu), mean(rmse_df_mu), mean(rmse_g_mu), mean(rmse_rf_mu))
    mean_ll[k,] <- c(mean(ll_dt), mean(ll_df), mean(ll_g), mean(ll_rf))

  cross_all <- cbind(rmse, ll)
  cross_mean <- cbind(mean_rmse, mean_ll)
  save(cross_all, file = "mcycle_cross_all.rda")
  save(cross_mean, file = "mcycle_cross_mean.rda")

}
@

\begin{figure}[t!]
\begin{subfigure}{0.6\textwidth}
<<mcycle_cross_rmse_mean, fig=TRUE, echo=FALSE>>=
# parallel boxplots RMSE
box_rmse <- cross_mean[, c("rmse_disttree", "rmse_randomForest", "rmse_distforest", "rmse_gamlss")]
colnames(box_rmse) <- c("disttree", "randomForest", "distforest", "gamlss")
boxplot.matrix(box_rmse, ylab = "RMSE")
@
\end{subfigure}
\hspace{-1cm}
\begin{subfigure}{0.6\textwidth}
<<mcycle_cross_ll_mean, fig=TRUE, echo=FALSE>>=
# parallel boxplots log-likelihood
box_ll <- cross_mean[, c("ll_disttree", "ll_randomForest", "ll_distforest", "ll_gamlss")]
#box_ll <- box_ll[box_ll[,1]>-400,]
#box_ll <- box_ll[box_ll[,2]>-400,]
box_ll <- box_ll[box_ll[,"ll_gamlss"]>-70,]
#box_ll <- box_ll[box_ll[,4]>-400,]
colnames(box_ll) <- c("disttree", "randomForest", "distforest", "gamlss")
boxplot.matrix(box_ll, ylab = "log-likelihood")
@
\end{subfigure}
%\center
\caption{\label{plot_cross_mean}10-times-10 cross validation on the mcycle data, averages (>-70):\\ \code{disttree} vs. \code{distforest} vs. \code{gamlss} vs. \code{randomForest}}
\end{figure}

%\begin{figure}[t!]
%\begin{subfigure}{0.6\textwidth}
%<<mcycle_cross_rmse_all, fig=TRUE, echo=FALSE>>=
%# parallel boxplots RMSE
%box_rmse <- cross_all[, c("rmse_disttree", "rmse_randomForest", "rmse_distforest", %"rmse_gamlss")]
%colnames(box_rmse) <- c("disttree", "randomForest", "distforest", "gamlss")
%boxplot.matrix(box_rmse, ylab = "RMSE")
%@
%%\caption{\label{plot_boot_rmse}Bootstraps: RMSE of the location parameter, \code{dist%tree%} vs. \code{distforest} vs. \code{gamlss} vs. \code{randomForest}}
%\end{subfigure}
%\hspace{-1cm}
%\begin{subfigure}{0.6\textwidth}
%<<mcycle_cross_ll_all, fig=TRUE, echo=FALSE>>=
%# parallel boxplots log-likelihood
%box_ll <- cross_all[, c("ll_disttree", "ll_randomForest", "ll_distforest", "ll_gamlss"%)]
%#box_ll <- box_ll[box_ll[,1]>-400,]
%#box_ll <- box_ll[box_ll[,2]>-400,]
%box_ll <- box_ll[box_ll[,"ll_gamlss"]>-100,]
%#box_ll <- box_ll[box_ll[,4]>-400,]
%colnames(box_ll) <- c("disttree", "randomForest", "distforest", "gamlss")
%boxplot.matrix(box_ll, ylab = "log-likelihood")
%@
%\end{subfigure}
%%\center
%\caption{\label{plot_cross_ll_all}mcycle 10-times-10 Cross validation (>-100): %\code{disttree} vs. \code{distforest} vs. \code{gamlss} vs. \code{randomForest}}
%\end{figure}

@@

>>>>

%% packages
\usepackage{amstext}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{thumbpdf}
\usepackage{rotating}
\usepackage{caption}
\captionsetup{format=hang}
\usepackage{subcaption}
\usepackage{Sweave}
\usepackage{enumitem}
\usepackage{tikz}
\usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}
%\usepackage[pagebackref=true]{hyperref}
%\usepackage{doi}

@@

\SweaveOpts{eval = TRUE}
\section{Introduction}

@@

All these parametric models work very well as long as only additive effects appear in the considered data. A different type of model that can capture non-additive effects as well are regression trees. Here the basic idea is to split the data set into more homogeneous subgroups based on information provided by the explanatory variables and to fit a model/response to each terminal node. One problem that might appear when applying tree algorithms is that they impose steps and abrupt changes on the model, even in situations where effects are rather smooth, as it can be seen in the top left plot of Figure \ref{Devel_treeforest}. One way how to tackle this problem is to not just consider one tree but an ensemble of trees. This is the idea of random forests. A set of trees is built, each on a slightly different data set which can either be a bootstrap or a subsample of the original data set. Then the results of all trees are combined to get a final model which unlike a tree model c!
 an also deal with smooth effects in a reasonable way.\\ 
In distributional trees and forests we now want to combine all these ideas from different fields of statistical modeling in order to benefit from all their advantages.\\
\\
In the past the combination of parametric data models and algorithmic tree models has gained more and more interest since it allows for a more flexible and clear structure. \\
Algorithms that only fit constants to each final node and therefore only model the mean of the response variable (such as CART, Breiman 1984) might lead to very large trees which complicates analyzing and interpreting. Algorithms such as GUIDE (Loh 2002), CRUISE (Kim and Loh 2001) and LOTUS (Chan and Loh 2004) were among the first to provide trees with parametric models in each terminal node. The MOB algorithm (Zeileis, Hothorn and Hornik 2008), an algorithm for model-based recursive partitioning, extended this idea by fitting parametric models to each node and using the gained information of the fitted model to create further splits in inner nodes. Using this algorithm as a framework to build trees LMs and GLMs can be fitted in the nodes of the trees. However, up to now trees which only model the mean of the response variable in its nodes are still more common and widely used.\\
Random forests (\cite{breiman2001random}) applying the CART algorithm to build the trees and cforest applying the ctree algorithm (\cite{hothorn2015partykit}, \cite{hothorn2006unbiased}) have used this idea of combining an ensemble of trees to get smoother effects of the covariates after combining or averaging over the parameters of the trees.\\
\\
But as mentioned before, until now the above listed development in the field of probabilistic modeling has not yet been fully integrated in the idea of algorithmic tree and forest models. Probabilistic models are rarely used in combination with tree algorithms and more flexible models such as GAMLSSs have not yet been applied as models in the nodes of a tree.\\
\\
Therefore, our main goal is to embed the progress made in the field of probabilistic modeling in the idea of regression trees and forests. In particular, more flexible models such as GAMLSSs should be fitted in the nodes of a tree in order to capture location, scale and shape as well as censoring, tail behavior etc. while non-additive effects or interactions of the explanatory variables can be detected automatically by the splitting algorithm used to build the tree. In that way, no variable selection or predefining of variable interactions is necessary and a whole distribution is specified in each node which leads to a wide range of inference methods that can be applied.\\
These (future) steps are illustrated in Figure \ref{Devel_treeforest}, embedded in the development made in the field of tree and forest algorithms.\\

@@

@@

\SweaveOpts{eval = TRUE}
\newpage
\newpage
\section{Rain}
In this application we analyze precipitation data meassured over 28 years at 95 (?) stations all over Tyrol and Vorarlberg. The considered data is provided by ..... and consists of the meassured total precipiation within 24 hours of each day in July from 1985 until 2012, both included. Additionally, numerical forecasts from the ECMWF are used as covariates. These forecasts are ensemble predictions consisting of ... individual forecasts based on perturbed initiall conditions.\\
In particular, 14 basic forecast variables are considered with up to 12 variations of these variables being included as covariates. Since each forecast variable consists of an ensemble of individual predictions the possible variations include the mean, the standarad deviation, the minimum and the maximum over all these individual predictions. Another variation refers to the time periods which can be over 24 hours or over time windows of 6 hours. Another possible option is to consider the mean over 24 hours for each individual prediction and then take the mean over the whole ensemble. In the following table listing all covariates this last option is referred to as the ensemble mean of mean.\\
The response variable total precicipitation is power transformed by a factor of 1.6. This is done due to experience and knowledge about possible distributions of precipitation meassurements. The resulting transformed variable is expected to be normally distributed and left censored at censoring point 0.

\newpage
The ensemble forecast for the total precipitation is represented by 12 covariates: 
\begin{itemize}[noitemsep]
\item ensemble mean of the sum of the total precipitation from 6--30 UTC
\item ensemble standard deviation of the sum of the total precipitation from 6--30 UTC
\item ensemble minimum of the sum of the total precipitation from 6--30 UTC
\item ensemble maximum of the sum of the total precipitation from 6--30 UTC
\item ensemble mean of the sum of the total precipitation from 6--12 UTC, from 12--18 UTC, from 18--24 UTC and from 24--30 UTC
\item ensemble standard deviation of the sum of the total precipitation from 6--12 UTC, from 12--18 UTC, from 18--24 UTC and from 24--30 UTC
\end{itemize}

The preciptable water is represented by 6 covariates:
\begin{itemize}[noitemsep]
\item the ensemble mean of the mean of preciptable water over the time period 6--30 UTC 
\item the ensemble mean of the minimum of preciptable water over the time period 6--30 UTC 
\item the ensemble mean of the maximum of preciptable water over the time period 6--30 UTC 
\item the ensemble standard deviation of the mean of preciptable water over the time period 6--30 UTC 
\item the ensemble standard deviation of the minimum of preciptable water over the time period 6--30 UTC 
\item the ensemble standard deviation of the maximum of preciptable water over the time period 6--30 UTC 
\end{itemize}

\newpage
\begin{tabular}{| l | c | l |}
\hline
basic covariates & nr & variations\\
\hline
\textbf{tp{\_}pow}: total precipitation    & 12 & ensemble mean, ensemble std dev, \\
\hspace*{\fill} power transformed (by 1.6) &    & ensemble min, ensemble max\\
\textbf{cape{\_}pow}: convective           &    & (all for 6--30 UTC)\\
\hspace*{0.4cm} available potential energy &    & ensemble mean: 6--12, 12--18,\\
\hspace*{\fill}  power transformed (by 1.6)&    & \hspace{2.5cm} 18--24, 24--30 UTC\\
\hspace*{\fill}                            &    & ensemble std dev: 6--12, 12--18,\\
\hspace*{\fill}                            &    & \hspace{2.85cm}  18--24, 24--30 UTC\\
\hline
\textbf{dswrf}: downwards short ware & 6 & ensemble mean of mean, \\
\hspace*{\fill} radiation flux (sunshine) &     & ensemble mean of min,\\
\textbf{msl}: mean sea level pressure  &     &  ensemble mean of max,\\
\textbf{pwat}: preciptable water &   & ensemble std dev of mean,\\
\textbf{tmax}: 2m maximum        &   & ensemble std dev of min,\\
\hspace*{1.15cm} temperature  &   & ensemble std dev of max,\\
\textbf{tcolc}: total column-integrated &   & (all for 6--30 UTC)\\
\hspace*{1.05cm} condensate     &   & \\
\textbf{t500}: temperature     &  & \\
\hspace*{1cm} on 500 hPA     &   & \\
\textbf{t700}: temperature     &  & \\
\hspace*{1cm} on 700 hPA     &   & \\
\textbf{t850}: temperature     &   & \\
\hspace{1cm} on 850 hPA     &   & \\
\hline
\textbf{tdiff500850}: temperature         & 3 & ensemble mean,\\
\hspace*{\fill} difference 500 to 850 hPA &   & ensemble min,\\
\textbf{tdiff500700}: temperature         &  & ensemble max\\
\hspace*{\fill} difference 500 to 700 hPA &   & (all for 6--30 UTC)\\
\textbf{tdiff700850}: temperature          &  & \\
\hspace*{\fill} difference 700 to 850 hPA  &   & \\
\hline
\textbf{msl{\_}diff}: mean sea level  & 1 & msl{\_}mean{\_}max - msl{\_}mean{\_}min\\
\hspace{1.65cm} difference            &   & (for 6--30 UTC)\\
\hline
\end{tabular}

\newpage

The following methods are applied on this data set:

\begin{tabular}{| l | c  | c  | c  | c |}
\hline
 & \textbf{mu link} & \textbf{sigma link} & \textbf{mu covariates} & \textbf{sigma covariates} \\
\hline
\textbf{disttree} & identity & log & all covariates & all covariates \\ 
\hline
\textbf{distforest} & identity & log & all covariates & all covariates \\
\hline
\textbf{gamlss} & identity & log & tppow{\_}mean, & tppow{\_}sprd,\\
& & & tppow{\_}max, & dswrf{\_}sprd{\_}mean, \\
& & & tppow{\_}mean1218, & tppow{\_}sprd1218,\\ 
& & & capepow{\_}mean1218, & capepow{\_}mean1218,\\
& & & dswrf{\_}mean{\_}mean, & tcolc{\_}sprd{\_}mean,\\
& & & tcolc{\_}mean{\_}mean, & tdiff500850{\_}mean\\
& & & pwat{\_}mean{\_}mean, & \\
& & & tdiff500850{\_}mean, & \\
& & & msl{\_}diff & \\
\hline
\textbf{gamboostLSS} & identity & log & all covariates & all covariates \\
\hline
\textbf{EMOS} & identity & log & tppow{\_}mean & tppow{\_}spread \\
\hline
\end{tabular}
\\
manuall selected covariates in gamlss\\
mention interactions:\\
tppow{\_}mean1218 $\cdot$ capepow{\_}mean1218,\\
tppow{\_}sprd1218 $\cdot$ capepow{\_}mean1218\\
\\
(advantage: no sepcification needed in disttree/distforest)

\SweaveOpts{eval = TRUE}
\newpage
\subsection{Station Axams}
one station \\
Axams, closest to Ibk\\
ca 800 observations\\

\subsubsection{Predictions}
Fit one model on the first 25 years.\\
predictions for the 26th, 27th and 28th year\\
plot predicted distributions together with obersvations for one day in July of 26th, 27th and 28th year  
<<echo=FALSE, results=hide>>=
#### Axams prediction 25 - 3
if(file.exists("rain_Axams_pred.rda")){
  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_Axams_pred.rda")
  load("rain_Axams_pred.rda")
} else {
  #source("~/svn/partykit/pkg/disttree/inst/draft/rain_axams_pred.R")
  source("rain_axams_pred.R")
  library("gamlss.cens")
  gen.cens("NO", type = "left")
  res <- rain_axams_pred(seedconst = 7, ntree = 100,
                         forest_mtry = 27,   # if frac == TRUE: nvar/3 = 30
                         gamboost_cvr = FALSE,
                         frac = FALSE)

  save(res, file = "rain_Axams_pred.rda")
}

# prepare data for plot of estimated density functions
pday <- 24  # 2 (hohe Beobachtung zu niedrig geschaetzt), 4, 15, evtl. auch 7, 8, 23 (eine 0-Beobachtung und 2 sehr aehnliche),

pdays <- if(pday<19) c(pday, pday + 31, pday + 61) else c(pday, pday + 30, pday + 61)
pdf <- predict(res$df, newdata = res$testdata[pdays,], type = "parameter")
df_mu <- pdf$mu
df_sigma <- pdf$sigma
df_exp <- pnorm(df_mu/df_sigma) * (df_mu + df_sigma * (dnorm(df_mu/df_sigma) / pnorm(df_mu/df_sigma)))
cbind(df_exp, res$testdata[pdays,"robs"])

# plot predicted distributions together with observations
#set.seed(res$call$seedconst)
set.seed(7)
x <- c(0.01, sort(runif(500,0.01,8)))
y1 <- crch::dcnorm(x, mean = df_mu[1], sd = df_sigma[1], left = 0)
y2 <- crch::dcnorm(x, mean = df_mu[2], sd = df_sigma[2], left = 0)
y3 <- crch::dcnorm(x, mean = df_mu[3], sd = df_sigma[3], left = 0)
dayending <- if(pday > 3) "th" else switch(pday, "1" = {dayending <- "st"}, "2" = {dayending <- "nd"}, "3" = {dayending <- "rd"})

# point mass (slightly shifted)
pm1 <- c(0.04, crch::dcnorm(-1, mean = df_mu[1], sd = df_sigma[1], left = 0))
pm2 <- c(-0.03, crch::dcnorm(-1, mean = df_mu[2], sd = df_sigma[2], left = 0))
pm3 <- c(-0.1, crch::dcnorm(-1, mean = df_mu[3], sd = df_sigma[3], left = 0))

# predictions
pred1 <- c(res$testdata[pdays,"robs"][1], crch::dcnorm(res$testdata[pdays,"robs"][1], mean = df_mu[1], sd = df_sigma[1], left = 0))
pred2 <- c(res$testdata[pdays,"robs"][2], crch::dcnorm(res$testdata[pdays,"robs"][2], mean = df_mu[2], sd = df_sigma[2], left = 0))
pred3 <- c(res$testdata[pdays,"robs"][3], crch::dcnorm(res$testdata[pdays,"robs"][3], mean = df_mu[3], sd = df_sigma[3], left = 0))

#legendheight
lh1 <- crch::dcnorm(0.01, mean = df_mu[1], sd = df_sigma[1], left = 0)
lh2 <- crch::dcnorm(0.01, mean = df_mu[2], sd = df_sigma[2], left = 0)
lh3 <- crch::dcnorm(0.01, mean = df_mu[3], sd = df_sigma[3], left = 0)
@
